---
title: "Data Cleaning Report"
author: "Arsénio Nhacolo"
institute: "Centro de Investigação em Saúde de Manhiça"
date: today
format:
  html:
    embed-resources: true
    self-contained-math: true
    css: aux/styles.css
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    code-link: false
    citation-location: document
bibliography: aux/references.bib
csl: aux/plos.csl
keep-md: false
execute:
  warning: false
  message: false
  echo: false
  cache: false
---

# Introduction {#sec-intro}

```{r}
#| label: setup
#| include: false
library(here)
library(readr)
library(dplyr)
library(knitr)

# Load datasets and error log
errors <- readr::read_rds(
    here::here("reports", "output", "cleaning", "cleaning_errors.rds")
)
raw_data <- readr::read_csv(
    here::here("data", "raw", "dat", "wsdd_data.csv"),
    na = c("", "NA", ".", "N/A"),
    show_col_types = FALSE
)
clean_data <- readr::read_rds(
    here::here("data", "processed", "dat", "clean_data.rds")
)

raw_count <- nrow(raw_data)
error_count <- nrow(errors)
clean_count <- nrow(clean_data)
```

This report covers the data cleaning and quality control process for the epidemiological study.

The raw dataset contains `r raw_count` records. During the quality control phase, a total of `r error_count` records were identified with logical inconsistencies. These records were excluded to ensure data integrity, leaving a final clean analytical cohort of `r clean_count` records.

# Discrepancy Log {#sec-discrepancy}

The identified errors are detailed in @tbl-cleaning-errors.

```{r}
#| label: tbl-cleaning-errors
#| tbl-cap: "Summary of detected inconsistencies in follow-up and event times."
#| echo: false
errors |>
    knitr::kable(
        col.names = c("Participant ID", "Variable", "Value", "Description"),
        align = "clll"
    )
```


# R Session Info

```{r}
#| label: session-info
#| echo: false
sessioninfo::session_info()
```
